---
output:
  github_document
---

<!---Note this will remove itself:-->

<!---
permalink: /teaching/chloropleths_101
title: Chloropleths + the malariaAtlas accessibility raster in one go!
categories:
  - howto
tags:
  - geo
  - R
toc: true
author_profile: false
--->

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Preamble

Here's a quick workshop to introduce some basic mapping techniques in R. We'll also have a look at some of the raster products released by the Malaria Atlas Project.

I'll show a mix of base graphics and `ggplot2`: while ggplot is often the way to go, it doesn't hurt to know about base graphics!

## Packages

Here are some bread and butter packages:

```{r pkgs, message=FALSE, warning=FALSE}
library(tidyverse)
theme_set(theme_bw()) # save myself some time later ...
library(sf)
library(terra) # for rasters
library(malariaAtlas)
```

## Polygons

Lots of spatial analysis relies on data associated with specific administrative units. For the purposes of this demonstration, let's retrieve admin level 1s in Nigeria.

There's lots of packages we can grab these sort of data from, e.g. `rnaturalearth` and `worlddatr`. Here's how you can do it with the MAP package:

```{r message=FALSE, warning=FALSE}
nigeria <- malariaAtlas::getShp(country = "Nigeria", 
                                admin_level = "admin1")

# in base graphics:
# the st_geometry() picks out the geometry from the simple feature collection
plot(st_geometry(nigeria))

# in ggplot
ggplot() +
  geom_sf(data = nigeria)
```

Have a look at `nigeria`: it's a simple feature collection that looks a bit like a data.frame. However, each row has geometry associated with it, as well as regular columns:

```{r}
head(nigeria)

# this gives us a data.frame without any geometry
nigeria %>%
  st_drop_geometry() %>%
  head()

# we can pick out rows just like a data frame:
plot(st_geometry(nigeria[1,]))
```

## Our first chloropleth

Let's have a first go at a chloropleth map ... in this example, I map the first letter of the state name:

```{r}
nigeria <- mutate(nigeria,
                  first_letter = substr(name_1, 1, 1) %>% factor())

ggplot() +
  geom_sf(data = nigeria,
          aes(fill = first_letter))
```

That's way too many colours! 
How about we just look at the five most common letters? And go for some groovy colours?

```{r}
most_common_letters <- nigeria %>%
  st_drop_geometry() %>%
  group_by(first_letter) %>%
  summarise(n = n()) %>%
  arrange(n) %>%
  tail(n = 5) %>%
  dplyr::select(first_letter) %>%
  unlist()

ggplot() +
  geom_sf(data = nigeria, fill = NA) + # give us all of the states first
  geom_sf(data = nigeria %>% 
            filter(first_letter %in% most_common_letters),
          aes(fill = first_letter)) + # now filled in states
  scale_fill_viridis_d(option = "A") +
  xlab("Longitude") +
  ylab("Latitude") +
  labs(title = "Nigerian states with the most common first letter")
```

Great! We've made some chloropleth maps to visualise discrete data. 

Now, let's download some raster data and have a look at it.

## Global accessibility map

Weiss et al. published a very useful [global map of travel time to cities](https://doi.org/10.1038/nature25181) in 2018, based on the world as in 2015. Have a read of the paper: the ultimate map of travel time to cities is a derived output from a global *friction surface* of travel time, where the value of each pixel is  the amount of time it takes to traverse it given the landscape and transport infrastructure.

Amelia Bertozzi-Villa has a [*medium* post](https://medium.com/@abertozz/mapping-travel-times-with-malariaatlas-and-friction-surfaces-f4960f584f08) describing how to use the friction surface for an arbitrary set of reference points. Using healthcare centres as reference points, Weiss et al. subsequently used the friction surface to examine [access to healthcare](https://doi.org/10.1038/s41591-020-1059-1). 

Here, we'll take a raster of the derived output, the estimated travel time to the nearest city of population greater than 50,000 people.

```{r message=FALSE}
# download the travel time raster:
#malariaAtlas::listRaster() # run this to check which rasters are avail

travel_time <- malariaAtlas::getRaster(dataset_id = "Accessibility__201501_Global_Travel_Time_to_Cities",
                        shp = nigeria)

travel_time
```

Our freshly downloaded `travel_time` is a `SpatRaster`, the raster format as represented in the `terra` package. A *raster* is a gridded dataset of values associated with two-dimensional coordinates (i.e., longitude and latitude), with a specific *resolution*, *extent*, and [*coordinate reference system*](https://en.wikipedia.org/wiki/Spatial_reference_system).

Let's have a look at our raster:

```{r}
# base graphics:
plot(travel_time, main = "Travel time to cities in Nigeria")
```

If we wanted to capture *accessibility*, we could invert travel time and rescale:

```{r}
access <- 1/(travel_time + 1e3)
ran <- range(values(access), na.rm=TRUE)
values(access) <- (values(access) - ran[1]) / 
  (ran[2] - ran[1])

plot(access, main = "Access to cities in Nigeria")
```

And in ggplot:

```{r}
# we'll need a bit of run-up:
access_df <- cbind(xyFromCell(access,
                              cells(access)),
                   as.data.frame(access))

ggplot() +
  # including the polygons sorts out the aspect of the raster
  geom_sf(data = nigeria) + 
  # access_df needs to be a data.frame here:
  geom_tile(data = access_df,
              aes(x = x, y = y, fill = `Travel Time to Cities`)) +
  scale_fill_viridis_c("Access")
```

## Bring it all together: chloropleth of accessibility

Let's make a chloropleth map of a continuous variable using our new accessibility raster. *Of course, I don't suggest that our accessibility raster is the best way to capture state-level "accessibility" in Nigeria - this map is for the purposes of demonstration.*

We first use `terra::extract()` to grab the mean of our accessibility surface for each state. The `na.rm = TRUE` excludes NAs from our mean calculation, and the `bind = TRUE` asks for the output to be bound to the polygons in the second argument, `nigeria`.

```{r}
access_states <- terra::extract(access, nigeria, mean, na.rm = TRUE, bind = TRUE) %>%
  # extract gives me a SpatVector .. convert back to sf:
  st_as_sf()

ggplot() +
  geom_sf(data = access_states, 
          aes(fill = Travel.Time.to.Cities)) +
  scale_fill_viridis_c("Mean access")
```

I highly recommend checking out Paula Moraga's 2023 textbook [*Spatial Statistics or Data Science: Theory and Practice with R*](https://www.paulamoraga.com/book-spatial/index.html). She goes through lots of what I've touched on here in more detail but it's all manageable for beginners!





